### 2018-01-05 ident/docs/ident_manuscript.tex
complete intro with outline

\begin{document}
\section{Introduction:}	
The dynamic response of any system can be modeled as a set of ordinary differential equations (ode) (Equation \ref{eq:output}a) in the state variables ($\mathbf{x}$). The changes in the state variables are expressed as a function of the state variables, the system parameters ($\theta$) and the inputs to the system ($u$).	
\begin{subequations}\label{eq:output}
	\begin{align}
	\dot{\mathbf{x}} = g(\mathbf{x}, \theta, u)\\
	\mathbf{y} = h(\mathbf{x},\mu,u)
	\end{align}
\end{subequations}
When all states are not observable, Equation (\ref{eq:output}a) is usually augmented with a system that defines the relationship between the observable output variables ($\mathbf{y}$) and the state variables ($\mathbf{x}$) as in Equation (\ref{eq:output}b). The parameters used to establish this relationship (Equation \ref{eq:output}b), $\mu$, may or may not include system parameters $\theta$ defined in Equation (\ref{eq:output}a).

Typically, a nonlinear programming formulation with the objective of minimizing the least square error between the measured ($y_{kl}^*$) and modeled ($y_{kl}$) outputs over the time course ($l = {1, ..., d}$) for which experimental data is collected (Equation \ref{eq:minnlp}) (Raue et al., and other parameter estimation/identifiability papers), is used to estimate parameters $\theta$ and $\mu$ in Equation \ref{eq:output}.
\begin{align}\label{eq:minnlp}
\chi^2(\theta,\mu) = \sum_{k=1}^{m}\sum_{l=1}^{d}\left(\frac{y_{kl}^*-y_{kl}}{\sigma_{kl}^*}\right)^2
\end{align}
The difference between the data and the model estimate for each output, and at each time point, is weighted by the variance in the experimental data $\sigma_{kl}^*$ for that corresponding variable and time point.
The minimization problem is solved subject to constraints on the state and output variables expressed in Equation \ref{eq:output}, as well as bounds on the estimated parameter values. The ability to determine a unique solution to the parameter estimation problem is however governed by the identifiability of the parameters in the model \parencite{McLean2012}.

The identifiability of parameters in nonlinear models can be classified into two categories: structural (or a priori) and practical (or posterior) identifiability. 	
Any system (Equation \ref{eq:output}) is said to be structurally identifiable if, for an input-output mapping defined by $\mathbf{y} = \Phi(\mu,u)$ for at least one input function $u$, any two values of parameters $\mathbf{\mu}_1$ and $\mathbf{\mu}_2$ satisfy the relationship in Equation (\ref{eq:stident}) below.
\begin{align}\label{eq:stident}
\Phi(\mathbf{\mu}_1,u) = \Phi(\mathbf{\mu}_2,u) \iff \mathbf{\mu}_1 = \mathbf{\mu}_2
\end{align}
Accordingly, any system that has an infinite number of solutions to the parameter estimation problem for all input functions is said to structurally non-identifiable. Thus, the structural identifiability of parameters in a dynamic model helps establish the presence or absence of a relationship between the unobservable system states and the observable system outputs. Accordingly, the effect of model structure and parameterization on the ability to infer true parameter values from experimental data is determined by the structural identifiability of the parameter.

@@

The dynamics of metabolic networks can also be represented by Equation (\ref{eq:output}) wherein the metabolite concentrations, $\mathbf{x}$, are state variables. The changes in the metabolite concentrations ($\mathbf{x}$) are expressed as a function of reaction rates dependent on enzyme kinetics, which are in turn expressed as nonlinear functions of the states ($\mathbf{x}$), kinetic parameters ($\theta$) and other input variables ($u$).

Algorithms have been extensively developed for establishing structural identifiability of dynamic models of biological systems (IEEE Trans paper from 2007). Most of these algorithms use methods based on differential algebra (Glad and Ljung, 1994). More recent methods take a profile-likelihood-based approach (2012 Paper) to establish both structural and practical identifiability. However, these methods not only scale poorly with increases in size of the modeled system, but also require dynamic time course data of the observable variables of the system. While computational burden due to poor scalability can be partly addressed with the current increases in computational power, the ability to obtain dynamic data for establishing identifiability of parameters in kinetic models of metabolism still remains a challenge.

In this paper, we propose a methodology to establish practical identifiability for parameters in kinetic models of metabolism. We present a computer algebra-based method that can facilitate experimental design for estimating parameters separately for each individual reaction within a metabolic network based on available steady state experimental data. This enables us to address the twin issues of scalability and data availability. We illustrate the utility of this method by applying it for a small network of gluconeogenesis in \textit{E. coli} and demonstrating our ability to propose experiments that will facilitate parameter estimation for a kinetic model of this network. We also demonstrate the scalability of the proposed methodology to facilitate experimental design by applying it to a relatively larger metabolic network of the human red blood cell hepatocyte.

\paragraph{Outline:}
\begin{itemize}
	\item Importance of parameter values for estimating in vivo response using kinetic models of metabolism and consequently for metabolic network design using kinetic models of metabolism
	\item The need for parameter identifiability to determine unique and true parameter values from observed data
	\item Types and purpose of identifiability for parameters
	\item Methods for structural identifiability and existing methods for practical identifiability
	\item Lack of methods for practical identifiability and consequently experimental design (not covered?)
	\item work done in this paper for practical identifiability			
	\item scalability of computer algebra-based methods for structural identifiability (using CRNT to reduce networks to make structural identifiability scalable) (move to discussion - may be in discussion)
\end{itemize}

>>>>

\begin{document}
\paragraph{Outline:}
\begin{itemize}
	\item Importance of parameter values for estimating in vivo response using kinetic models of metabolism and consequently for metabolic network design using kinetic models of metabolism
	\item The need for parameter identifiability to determine unique and true parameter values from observed data
	\item Types and purpose of identifiability for parameters
	\item Methods for structural identifiability and existing methods for practical identifiability
	\item Lack of methods for practical identifiability and consequently experimental design (not covered?)
	\item work done in this paper for practical identifiability			
	\item scalability of computer algebra-based methods for structural identifiability (using CRNT to reduce networks to make structural identifiability scalable) (move to discussion - may be in discussion)
\end{itemize}
\section{Introduction:}
Kinetic models of metabolism can be used to study the dynamic characteristics of metabolic networks. In these models, ordinary differential equations (ode) are used to express the rate of change of metabolite concentrations ($x$) as a function of the reaction fluxes ($v$) in the metabolic network (Equation \ref{eq:kinstoich}). The matrix $\mathbf{S}$ in Equation (\ref{eq:kinstoich}a) defines the stoichiometric relationship between the fluxes and the concentrations of the metabolic network.
\begin{subequations}\label{eq:kinstoich}
	\begin{align}
	\dot{x} = \mathbf{S}v\\
	v = g(x, \theta, u)
	\end{align}
\end{subequations}
The expression for the nonlinear function ($g$) used to describe each reaction flux $v_i$ in a kinetic model (Equation \ref{eq:kinstoich}b) is dependent on the enzyme kinetic mechanism that is used to model the reaction (older kinetic modeling papers, reviews, Srinivasan et al., 2015). Accordingly, $g$ is a nonlinear function of the metabolite concentrations, enzyme kinetic parameters ($\theta$) and other input concentrations ($u$).

The ability to predict the steady state and dynamic responses of metabolic networks, under in vivo conditions, to different perturbations is dependent on the numerical values of the enzyme kinetic parameter values ($\theta$) in Equation (\ref{eq:kinstoich}). Analyzing the ability of a metabolic network to exhibit dynamic characteristics like multiple steady states and oscillations, irrespective of the structure of the network, is one example where parameter values might play a crucial role (Srinivasan et al., 2017?). The use of in vitro, or unreliable in vivo parameter estimates, reduces confidence in the model predicted behaviour. Consequently, the reduction in confidence hampers the use of these models to gain insight into the functioning of metabolic networks (ensemble modeling and ORACLE papers). The insights gained from the use of kinetic models are subsequently used to design changes to these metabolic networks to achieve various goals. These goals could either be to increase metabolite production for biosynthesis of different chemicals (kinetic model for BDO in ecoli, k-ecoli457, Srinivasan et al., 2017) or to find therapeutic targets that can cut-off metabolic pathways to kill cancer cells (new paper on using geneMCS for cancer cell lines). Hence, an increase in uncertainty in model predicted responses is also an obstacle for using the predicted responses as a basis for designing the metabolic networks to achieve these goals.

If all intracellular metabolite concentrations can be measured over a time course, and the corresponding data is available, a nonlinear programming problem can be formulated to estimate the enzyme kinetic parameters ($\theta$) in Equation (\ref{eq:kinstoich}). The minimization of least square error between the measured ($x^*$) and modeled ($x$) concentrations can be used as an objective function (Equation \ref{eq:lsqopt}a) for the optimization problem (Equation \ref{eq:lsqopt}), with fixed upper ($\theta_u$) and lower ($\theta_l$) bounds for the parameter values (Equation \ref{eq:lsqopt}b). The difference between the data and the model estimate for each output and at each time point in the objective function is weighted by the variance in the experimental data $\sigma_{kl}^*$ for that corresponding variable and time point.
\begin{subequations}\label{eq:lsqopt}
	\begin{align}
	\underset{\theta}{\mathrm{min}} &\sum_{k=1}^{m}\sum_{l=1}^{d}\left(\frac{x_{kl}^*-x_{kl}}{\sigma_{kl}^*}\right)^2\\
	&\theta_l \le \theta \le \theta_u
	\end{align}
\end{subequations}

However, not all metabolite concentrations used in the model (Equation \ref{eq:kinstoich}) can be measured. Additionally, measurable fluxes in the metabolic network also need to be included as part of the parameter estimation problem. In such scenarios, the parameter estimation problem is modified to suit a new system of equations shown below (Equation \ref{eq:output}). The new system of equations is obtained by augmenting the original system (Equation \ref{eq:kinstoich}) with Equation (\ref{eq:output}c) that models the relationship between the measurable metabolite concentrations and fluxes, represented by the variable $y$, and the unmeasured concentrations ($x$) that are used in the original model (Equation \ref{eq:kinstoich}) above. Equation (\ref{eq:output}) now has additional parameters $\mu$ that define this relationship and also need to be estimated.
\begin{subequations}\label{eq:output}
	\begin{align}
	\dot{x} = \mathbf{S}v\\
	v = g(x, y, \theta, u)\\
	\dot{y} = h(x, y, \mu, u)
	\end{align}
\end{subequations}
In systems identification, the measured concentrations and fluxes, $y$ in Equation (\ref{eq:output}c), are called output or observed variables, and the unmeasured concentrations, $x$ in Equation (\ref{eq:output}a-b), are called the state variables. 
For estimating both $\theta$ and $\mu$, the metabolite concentrations ($x$) in the optimization problem (Equation \ref{eq:lsqopt}) are substituted with the output variables $y$.

However, the ability to determine unique solutions to parameters $\theta$ and $\mu$ is governed by the identifiability of these parameters in the model \parencite{McLean2012}. The identifiability of parameters in nonlinear models can be classified into two categories: structural (or a priori) and practical (or posterior) identifiability. 	
Any system (Equation \ref{eq:output}) is said to be structurally identifiable if, for an input-output mapping defined by $\mathbf{y} = \Phi(\mu,u)$ for at least one input function $u$, any two values of parameters $\mathbf{\mu}_1$ and $\mathbf{\mu}_2$ satisfy the relationship in Equation (\ref{eq:stident}) below.
\begin{align}\label{eq:stident}
\Phi(\mathbf{\mu}_1,u) = \Phi(\mathbf{\mu}_2,u) \iff \mathbf{\mu}_1 = \mathbf{\mu}_2
\end{align}
Accordingly, any system that has an infinite number of solutions to the parameter estimation problem for all input functions is said to structurally non-identifiable. Thus, the structural identifiability of parameters in a dynamic model helps establish the presence or absence of a relationship between the unobservable state variables and the observable output variables. Accordingly, the effect of model structure and parameterization on the ability to infer true parameter values from experimental data is determined by the structural identifiability of the parameter.

@@

Algorithms have been extensively developed for establishing structural identifiability of dynamic models of biological systems (IEEE Trans paper from 2007). Most of these algorithms use methods based on differential algebra (Glad and Ljung, 1994). More recent methods take a profile-likelihood-based approach (2012 Paper) to establish both structural and practical identifiability. However, these methods scale poorly with increase in size of the modeled system. The aforementioned methods are also heavily dependent on the availability of dynamic time course data for all the output variables of the system.

Computational burden due to poor scalability can be partly addressed with the current increases in computational power, while current improvements in measuring technologies can enable one to obtain high frequency dynamic data, no scalable methodologies exist to utilize the abundantly available steady state data for parameter estimation and identifiability for large metabolic networks.

In this paper, we propose a methodology to establish practical identifiability for parameters in kinetic models of metabolism. We present a computer algebra-based method that can facilitate practical identifiability as well as experimental design for estimating parameters separately for each individual reaction within a metabolic network based on available steady state experimental data. This enables us to address the twin issues of scalability and data availability. We illustrate the utility of this method by applying it for a small network of gluconeogenesis in \textit{E. coli} and demonstrating our ability to propose experiments that will facilitate parameter estimation for a kinetic model of this network. We also demonstrate the scalability of the proposed methodology to facilitate experimental design by applying it to a relatively larger metabolic network of the human red blood cell hepatocyte.
